What is the most dedicated downstream for a selected upstream?
========================================================

Alexander Favorov  
-----------------------------------------------

### `r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r options(width=80,keep.source=TRUE)`
`r opts_chunk$set(echo=T,message=F,cache=F,warning=F)`
<!-- cache is off, we save all the necessary things in Rda -->

Both questions and much more questions of this kind (which gene behaves the most specifically like a selected one, etc..) based on the following consideration. Suppose we have any two sets, $A$ and $B$, and we have a pairwise relation measure $\left<a,b\right>$ for each of element $(a,b)$ of their Cartesian product $A x B$. Let's say the relation is similarity rather than distance, the higher it is, the more the elements in the pair are related. $A$ and $B$ could be the same. If a pair $(a,b)$ does not make a sense for the relation we describe, we can say that the measure is $\left<a,b\right>=0$. Now, for each selected $a \in A$ we can make a ranking of all the $b \in B$ elements based on the relation $\left<a,b\right>$. But, actually, the top of the ranking can be filles with $b$'s that do not specifically relate to the selected $a$, they just strongly interact with all or almost all the $a$'s. Sometimes, it makes sense (if we wnat to know what are strongest regulators for gene G?), but sometimes, we want to answer other question: what is the strongest specific regulator for gene G (let's call it the best friend of G). In our notation, $A$ is the set of genes, and $B$ is the set of regulators, and $ \left<a,b\right> $ is some regulatory strenght.

Let's think about ranks. Suppose we ranked all the $b$'s by $\left<a,b\right>$ for a selected $a$. The result is $Rank_a(b)= \# \left\{ \beta \in B : \left< a, \beta \right> \leqslant \left< a, b \right> \right\}$. To answer the 'best-friend' question, we are to calculate more complicated value. We want to raise up those $b$'s that are specifically similar to our $a$, so we first rank all $a$'s for all $b$'s by the $\left<a,b\right>$  values and then rank the ranks our selected $a$ get in the rankings for different $b$'s. The best rank $b$'s now are those that provide unusially high rank for our selcted $a$ in their ranking.

$$BFRank_a(b)=Rank_a(Rank_b(a))$$

Please note, the $\left<a,b\right>$ measure is not to be symmetric, the backwards value $\left<b,a\right>$ is never used and thus even not required to be defined.


```{r source, echo=FALSE}
if (!suppressWarnings(require('best.friends.of')))
{
	if (!suppressWarnings(require('devtools')))
	{
		source("http://bioconductor.org/biocLite.R")
		biocLite("devtools")
		library("devtools")
	}
	install_github('favorov/gene-friends/pkg')
	library(best.friends.of)
}
```

### Code and examples in R

#### Best Friends ranking

And now, let's prepare a simple example. We have five genes (A,B,C,D,E) and three TFs (TF1,TF2,TF3). The strength of the regulation is shown by the matrix. 

```{r}
genes<-10
gene.names<-LETTERS[seq( from = 1, to = genes )]
regulation=matrix(
	c(rep(1, genes),
		0.2, 0.2, 0.2, 0.2, 0.25, rep(0.2,genes-5),
		rep(1, genes),
		rep(1, genes),
		rep(1, genes),
		rep(1, genes),
		rep(1, genes),
		rep(1, genes),
		rep(1, genes),
		rep(1, genes)
    ),
	ncol=10,byrow=FALSE
)
TF.names<-c('TF1','TF2','TF3','TF4','TF5','TF6','TF7','TF8','TF9','TF10')
rownames(regulation)<-gene.names
colnames(regulation)<-TF.names
regulation
```

Applying the function to our example.

```{r}
ranklists<-RankByBestFriendOf(regulation)
ranklists
```
#### Best Friends ordering

```{r}
orderlists<-OrderByBestFriendOf(regulation)
orderlists
```

We want to see gene names rather than indices:

```{r}
TFlists<-apply(orderlists,2,function(col) TF.names[col])
TFlists
```


